Medical Concept Extraction

### 2012-12-16 paper/paper.tex
Starting on the data section.

\subsubsection{Porter Stemmer}
We already discussed the benefits of using entire words as features, but similar words come in many different forms. Motivated by generalization, we chose to use word stems as features and approached this feature extraction using three different stemming algorithms--two at the word-level and one at the sentence level (See 5.3.2). The addition of word stem features to our system improved our testing CRF F1 scores from 0.58 to 0.82.

Namely, we used the Porter Stemming Algorithm, as developed by Martin Porter.  This particular stemming algorithm is divided into a number of linear steps. Word letters are first replaced by symbols for vowels, V and consonants, C. The measure, m, of the word is computed as the number of repetitions of the pattern VC, and this is used to determine whether suffixes ought to be removed. Plural word forms, such as  sses $\rightarrow$ ss and removal of trailing s's, are then removed; followed by the removal of past participles; and finally terminal y's are replaced with i's. Subsequent steps handle different orderings of suffices by translating double suffices into a single suffix and then removing suffices under certain predefined conditions.

@@

Derived from the Porter Stemmer, the Lancaster Stemmer was used to provide a second word stem feature to our system. By constant, this stemming algorithm uses a table of rules that specific the removal or replacement of an ending. Endings are replaced to maintain properly-spelled words throughout the stemming process--this avoids the need for an additional recoding or partial matching phase.

The rule table looks for the following information about each word: Endings of one or more characters (stored in reverse order); an optional intact flag; a digit specifying the removal total;  an optional append string of one or more characters;  and a continuation symbol. The stemmer first inspects the final letter of a term and considers the first relevant rule in the table. If the rule conditions are not satisfied, another rule is considered, otherwise the rule is applied and the ending is removed or reformed. This process is repeated until a termination symbol is uncovered and the word is presumed to be in its stem form.

\subsubsection{Word Shape}
In the same spirit of our word length and MIT RegEx features, we decided to look at the high-level shapes of individual words for better generalization. More specifically, we encode words into new strings that give a sense of their capitalization and punctuation patterns as well as approximate length. We tested a variety of open-sourced word shape classification methods including two methods written by Dan Klein, one of which simply classifies words as numeric, strictly upper or lower cased, or mixed, and a second, more fine-grained classifier that attempts to group words into equivalence classes by collapsing sequences of the same time. Additionally classifiers developed by Christopher Manning seek to make less distinctions mid-sequence by sorting words based on length. Each word shape classifier result added an additional dimension to our feature space. By naively including all word shape alternatives as features, we were able to improve our testbed CRF F1 score from 0.82 to 0.85. After further testing, we settled on \begin{verbatim}______\end{verbatim} as the most promising word shape metric because \begin{verbatim}______.\end{verbatim}

@@

\subsubsection{WordNet Lemmatizer}
Unlike the Porter Stemmer and the Lancaster Stemmer, which consider words independent of context, the WordNet Lemmatizer requires words to be processed at the sentence level in order to extract their root form. The WordNet Lemmatizer uses the part of speech tags for individual words, which are assign via NLTK as previously mentioned, in conjunction with a subset of tags used for morphology to extract stems from each word. WordNet references a database of word stems, and uses two types of processes to try to convert input strings into a form present in the database. A list of inflectional endings based on the word's syntactic tag is referenced in search of a match that can be detached from the word. An exception list is also referenced in search for an inflected form. WordNet first checks for exceptions and then uses rules for detachment to transform the word--after each transformation the result is referenced against the stem database until a match is eventually found.

@@

Talk about the competition and where the data came from. How large the data size is, how much we trained on, etc.

>>>>

\subsubsection{Porter Stemmer}
We already discussed the benefits of using entire words as features, but similar words come in many different forms. Motivated by generalization, we chose to use word stems as features and approached this feature extraction using three different stemming algorithms--two at the word-level and one at the sentence level (See 5.3.2). Namely, we used the Porter Stemming Algorithm, as developed by Martin Porter.

TODO(SAM):

Our CRF F1 scores went from 0.58 to 0.82

@@

TODO(SAM):

\subsubsection{Word Shape}
In the same spirit of our word length and MIT RegEx features, we decided to look at the high-level shapes of individual words for better generalization. More specifically, we encode words into new strings that give a sense of their capitalization and punctuation patterns as well as approximate length. We tested a variety of open-sourced word shape classification methods including two methods written by Dan Klein, one of which simply classifies words as numeric, strictly upper or lower cased, or mixed, and a second, more fine-grained classifier that attempts to group words into equivalence classes by collapsing sequences of the same time. Additionally classifiers developed by Christopher Manning seek to make less distinctions mid-sequence by sorting words based on length. Each word shape classifier result added an additional dimension to our feature space. By naively including all word shape alternatives as features, we were able to improve our testbed CRF F1 score from 0.82 to 0.85. After further testing, we settled on \begin{verbatim}______\end{verbatim} as the most promising word shape metric because \begin{verbatim}______\end{verbatim}.
\begin{verbatim}
  ()_()
 ( ' ' )
   >.<
 (")_(")
\end{verbatim}

@@

\subsubsection{WordNet Lemmatizer}
TODO(SAM):

@@

The data set that we used train and test the system came from the i2b2 NLP challenge. As mentioned before, the data was already divided into test and training sets and every word was pre-labeled with the following classes: ``problem'', ``test'', ``treatment'', or ``none''.

The training data consisted of 170 medical records and the test data consisted of 256 medical records.  The overall size of the data was 10mb. The data all came from Boston-area hospitals and is subject to a data usage agreement.
